WHOOMP! (There It Is) \\ Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2015-07-07 ms.tex
Describe two observing runs

\begin{abstract}
Within the next few years, Advanced LIGO and Virgo should detect \acp{GW} from binary neutron star and neutron star\nobreakdashes--black hole mergers. These sources are also predicted to power a broad array of electromagnetic transients. Because the X\nobreakdashes-ray and optical signatures can be faint and fade rapidly, observing them hinges on rapidly inferring the sky location from the gravitational wave observations. MCMC methods for gravitational\nobreakdashes-wave parameter estimation can take hours or more. We introduce BAYESTAR, a rapid, Bayesian, non-MCMC sky localization algorithm that takes just seconds to produce probability sky maps that are comparable in accuracy to the full analysis. Prompt localizations from BAYESTAR will make it possible to search electromagnetic counterparts of compact binary mergers.
\end{abstract}

@@

The innermost integral over distance $r$ may be written as
%
\begin{align}\label{eq:distance-integral}
    \mathscr{F} &=
        \int_{\mathrlap{r_\mathrm{min}}}^{\mathrlap{r_\mathrm{max}}}
        \exp\left[-\frac{p^2}{r^2}\right]
        I_0\left[\frac{b}{r}\right] r^m dr \nonumber \\
    &=
        \int_{\mathrlap{r_\mathrm{min}}}^{\mathrlap{r_\mathrm{max}}}
        \exp\left[-\frac{p^2}{r^2}+\frac{b}{r}\right]
        \overline{I}_0\left[\frac{b}{r}\right] r^m dr
\end{align}
%
or, completing the square,
%
\begin{align}
    \mathscr{F} &= \exp\left[\frac{p^2}{{r_0}^2}\right]
        \int_{\mathrlap{r_\mathrm{min}}}^{\mathrlap{r_\mathrm{max}}}
        \exp\left[-\left(\frac{p}{r} - \frac{p}{r_0}\right)^2\right]
        \overline{I}_0\left[\frac{2p^2}{r r_0}\right] r^m dr \\
    &= \exp\left[\frac{p^2}{{r_0}^2}\right] \mathscr{G},
\end{align}
%
where
%
\begin{align}
    p^2 &= \frac{1}{2} r^2 \sum_i {\rho_i}^2 \label{eq:likelihood-p-factor} \\
    b &= r \left| \sum_i \hat\rho_i \rho_i e^{i \tilde{\gamma}_i}
        a_i^*(\tilde{\tau}_i) \right| \label{eq:likelihood-b-factor} \\
    r_0 &= 2 p^2 / b \\
    \overline{I}_0(x) &= \exp(-|x|) I_0(x).
\end{align}
%
The coefficients $p^2$ and $b$ are nonnegative and independent of distance. $p$ has a maximum value of
%
\begin{equation}
    p_\mathrm{max} = \sqrt{\frac{1}{2} \sum_i \left(\frac{r_{1,i}}{r}\right)^2}.
\end{equation}
%
The symbol $\overline{I}_0$ denotes an exponentially scaled Bessel function. In the limit of large argument, $I_0(|x|) \sim \exp(|x|) / \sqrt{2 \pi |x|}$ \citep{Olver:2010:NHMF,NIST:DLMF}\footnote{\url{http://dlmf.nist.gov/10.40.E1}}. The scaled Bessel is useful for evaluation on a computer because it has a relatively small range $(0, 1]$ and varies slowly in proportion to $x^{1/2}$.

@@

\begin{figure}
    \includegraphics[width=\columnwidth]{adaptive_mesh}
    \caption{\label{fig:adaptive-mesh}An example multi\nobreakdashes-resultion \ac{HEALPix} mesh arising from the \ac{BAYESTAR} sampling scheme. This is event 18951 from \cite{FirstTwoYears}.}
\end{figure}

@@

We have completed our description of the \ac{BAYESTAR} algorithm. In the following chapter, we will describe the first results with \ac{BAYESTAR}, a detailed case study of the sky localization capabilities of the earliest configurations of Advanced \ac{LIGO} and Virgo. Because it is typically three orders of magnitude faster than the conventional \ac{MCMC} analysis, for the first time we are able to generate detailed sky maps for thousands of simulated \ac{GW} sources. With such a large sample, we can for the first time unambiguously describe areas and morphologies that should be typical of networks of just two detectors. This case study will be the subject of the next chapter.

@@

Several features are apparent. First, at any number of threads, the two configurations have similar run times, although the 2016 events contain a subpopulation of outliers that take about 2.5 times as long as the 2015 events. These are probably due to taking one of the more expensive code branches in the distance integral. Second, the run times decrease proportionally to the number of threads. A slight flattening at 32 threads may be due to increasing use of hyperthreading. Based on experiences running \ac{BAYESTAR} on the 32\nobreakdashes-core (64 threads with hyperthreading) cluster login machine, we expect the almost ideal parallel speedup to continue on machines with more threads. With just one thread, the \ac{BAYESTAR} analysis takes 500\nobreakdashes--1500~s, already orders of magnitude faster than the full parameter estimation. With 32 threads, \ac{BAYESTAR} takes just 20\nobreakdashes--60~s. This latency is comparable to the other stages (data aggregation, trigger generation, alert distribution) in the real\nobreakdashes-time \ac{BNS} analysis. The 32\nobreakdashes-thread configuration is representative of how \ac{BAYESTAR} might be deployed in early Advanced \ac{LIGO}.\footnote{\ac{BAYESTAR} has been successfully ported to the Intel's Many Integrated Core~(MIC) architecture and has been tested in a 500 thread configuration on a system with dual Intel Xeon Phi 5110P coprocessors.}

>>>>

\begin{abstract}
Within the next few years, Advanced LIGO and Virgo should detect \acp{GW} from binary neutron star and neutron star\nobreakdashes--black hole mergers. These sources are also predicted to power a broad array of electromagnetic transients. Because the X\nobreakdashes-ray and optical signatures can be faint and fade rapidly, observing them hinges on rapidly inferring the sky location from the gravitational wave observations. \ac{MCMC} methods for gravitational\nobreakdashes-wave parameter estimation can take hours or more. We introduce BAYESTAR, a rapid, Bayesian, non-\ac{MCMC} sky localization algorithm that takes just seconds to produce probability sky maps that are comparable in accuracy to the full analysis. Prompt localizations from BAYESTAR will make it possible to search electromagnetic counterparts of compact binary mergers.
\end{abstract}

@@

The innermost integral over distance\footnote{The distance integral is now performed differently from what we initially described in \cite{leo-singer-thesis}; the method described in the present work is about an order of magnitude faster.} $r$ may be written as
%
\begin{align}\label{eq:distance-integral}
    \mathscr{F} &=
        \int_{\mathrlap{r_\mathrm{min}}}^{\mathrlap{r_\mathrm{max}}}
        \exp\left[-\frac{p^2}{r^2}\right]
        I_0\left[\frac{b}{r}\right] r^m dr \nonumber \\
    &=
        \int_{\mathrlap{r_\mathrm{min}}}^{\mathrlap{r_\mathrm{max}}}
        \exp\left[-\frac{p^2}{r^2}+\frac{b}{r}\right]
        \overline{I}_0\left[\frac{b}{r}\right] r^m dr
\end{align}
%
or, completing the square,
%
\begin{align}
    \mathscr{F} &= \exp\left[\frac{p^2}{{r_0}^2}\right]
        \int_{\mathrlap{r_\mathrm{min}}}^{\mathrlap{r_\mathrm{max}}}
        \exp\left[-\left(\frac{p}{r} - \frac{p}{r_0}\right)^2\right]
        \overline{I}_0\left[\frac{2p^2}{r r_0}\right] r^m dr \\
    &= \exp\left[\frac{p^2}{{r_0}^2}\right] \mathscr{G},
\end{align}
%
where
%
\begin{align}
    p^2 &= \frac{1}{2} r^2 \sum_i {\rho_i}^2 \label{eq:likelihood-p-factor} \\
    b &= r \left| \sum_i \hat\rho_i \rho_i e^{i \tilde{\gamma}_i}
        a_i^*(\tilde{\tau}_i) \right| \label{eq:likelihood-b-factor} \\
    r_0 &= 2 p^2 / b \\
    \overline{I}_0(x) &= \exp(-|x|) I_0(x).
\end{align}
%
The coefficients $p^2$ and $b$ are nonnegative and independent of distance. $p$ has a maximum value of
%
\begin{equation}
    p_\mathrm{max} = \sqrt{\frac{1}{2} \sum_i \left(\frac{r_{1,i}}{r}\right)^2}.
\end{equation}
%
The symbol $\overline{I}_0$ denotes an exponentially scaled Bessel function. In the limit of large argument, $I_0(|x|) \sim \exp(|x|) / \sqrt{2 \pi |x|}$ \citep{Olver:2010:NHMF,NIST:DLMF}\footnote{\url{http://dlmf.nist.gov/10.40.E1}}. The scaled Bessel is useful for evaluation on a computer because it has a relatively small range $(0, 1]$ and varies slowly in proportion to $x^{1/2}$.

@@

\begin{figure}
    \includegraphics[width=\columnwidth]{adaptive_mesh}
    \caption{\label{fig:adaptive-mesh}An example multi\nobreakdashes-resultion \ac{HEALPix} mesh arising from the \ac{BAYESTAR} sampling scheme (plotted in a cylindrical projection). This is event 18951 from \cite{FirstTwoYears}.}
\end{figure}

@@

We have completed our description of the \ac{BAYESTAR} algorithm. In \cite{FirstTwoYears}, the authors presented a comprehensive and astrophysically realistic sample of simulated \ac{BNS} mergers. We focused on the first two planned Advanced LIGO and Virgo observing runs as desribed in \cite{LIGOObservingScenarios}. That work presented a catalog of 500 sky localizations from \ac{BAYESTAR} and LALINFERENCE, and dealt with the quantitave position reconstruction accuracy as well as the qualitative sky morphologies. In the present work, we will use the same data set, but instead focus on demonstrating the correctness and performance of the \ac{BAYESTAR} algorithm.

\subsection{Observing scenarios}

To review the assumptions made in \cite{FirstTwoYears}, the two scenarios are:

\paragraph*{2015} The first Advanced \acs{LIGO} observing run, or ``O1'', scheduled to start in September 2015 and continue for three months. There are only two detectors participating in this run: \acs{LIGO} Hanford (``H'') and \acs{LIGO} Livingston (``L''). Both detectors are expected to operate with a direction\nobreakdashes-averaged \ac{BNS} merger range of 40\nobreakdashes--80\,Mpc (though ongoing Advanced \acs{LIGO} commissioning suggests that the higher end of this range will be achieved). As a result of having only two detectors, most localizations are long, thin arcs a few degrees wide and tens to hundreds of degrees long. The median 90\% credible area is about 600\,deg$^2$.

\paragraph*{2016} The second observing run, ``O1'', with the two between Advanced \acs{LIGO} detectors, upgraded to a \ac{BNS} range of 80\nobreakdashes--120\,Mpc, operated jointly with the newly commissioned Advanced Virgo detector (``V''), operating at a range of 20\nobreakdashes--60\,Mpc. The run is envisioned as lasting for six months in 2016\nobreakdashes--2017. The detectors are assumed to have random and independent 80\% duty cycles. Consequently, all three detectors (``HLV'') are in science mode about half of the time, with the remaining time divided roughly equally between each of the possible pairs (``HL'', ``HV'', or ``LV'') and one or fewer detectors (at least two \ac{GW} facilities are required for a detection). Virgo's range is assumed to be somewhat less than \acs{LIGO}'s because its commissioning timetable is about a year behind. Although the simulated signals are generally to weak in Virgo to trigger the matched\nobreakdashes-filter pipeline and contribute to \emph{detection}, even these sub\nobreakdashes-threshold signals aid in position reconstruction with LALINFERENCE by lifting degeneracies. As a result, the median 90\% credible area decreases to about 200\,deg$^2$.

\subsection{Self-consistency of credible levels}

One might naively expect that self\nobreakdashes-consistency would require the 90\% confidence area and searched area histograms to intersect at 90\% of detections, but this is not generally required because the posteriors of different events have widely different dimensions. However, it is true that 90\% of sources should be found within their respective 90\% confidence contours. This can be formalized into a graphical self\nobreakdashes-consistency test called a probability\nobreakdashes--probability ($P$\nobreakdashes--$P$) plot (see \citealt{SiderySkyLocalizationComparison} for applications in \ac{GW} parameter estimation). For each event, one follows the water\nobreakdashes-filling procedure to walk outward from the most probable pixel until one reaches the pixel that contains the true sky location, noting the total probability $p$ that has been traversed.

A cumulative histogram of $p$ should be diagonal, within a binomial confidence band. It is already well established that LALINFERENCE\_NEST/MCMC/BAMBI satisfy the $P$\nobreakdashes--$P$ plot test when deployed with accurate templates and reasonable priors. We found at first that \ac{BAYESTAR}'s $P$\nobreakdashes--$P$ plots tended to sag below the diagonal, indicating that though the accuracy (i.e., searched area) was comparable to the stochastic samplers, the \emph{precision} was overstated with confidence intervals that were only about 70\% of the correct area. This was rectified by pre\nobreakdashes-scaling the \acp{SNR} from GSTLAL\_INSPIRAL by a factor of 0.83 prior to running \ac{BAYESTAR}. This correction factor suggests that, for example, an \ac{SNR} 10 trigger from GSTLAL\_INSPIRAL has the \emph{effective information content} of an \ac{SNR} 8.3 signal. The missing information may be due to losses from the discreteness of the template bank, from the \ac{SVD}, from mismatch between the matched\nobreakdashes-filter templates and the simulated signals, from the small but nonzero correlations between masses and intrinsic parameters, or from elsewhere within the detection pipeline. The correction is hard\nobreakdashes-coded into the rapid localization. With it, the $P$\nobreakdashes--$P$ plots are diagonalized without negatively affecting the searched area (see Figure~\ref{fig:pp}).

\begin{figure}
    \begin{center}
        \includegraphics{pp}
    \end{center}
    \caption[\acs{BAYESTAR} $P$\nobreakdashes--$P$ plots]{\label{fig:pp}$P$\nobreakdashes--$P$ plots for \ac{BAYESTAR} localizations in the 2015 and 2016 configurations. The gray lozenge around the diagonal is a target 95\% confidence band derived from a binomial distribution.}
\end{figure}

@@

Several features are apparent. First, at any number of threads, the two configurations have similar run times, although the 2016 events contain a subpopulation of outliers that take about 2.5 times as long as the 2015 events. These are probably due to taking one of the more expensive code branches in the distance integral interpolation. Second, the run times decrease proportionally to the number of threads. A slight flattening at 32 threads may be due to increasing use of hyperthreading. Based on experiences running \ac{BAYESTAR} on the 32\nobreakdashes-core (64 threads with hyperthreading) cluster login machine, we expect the almost ideal parallel speedup to continue on machines with more threads. With just one thread, the \ac{BAYESTAR} analysis takes 100\nobreakdashes--400~s, already orders of magnitude faster than the full parameter estimation. With 32 threads, \ac{BAYESTAR} takes just 20\nobreakdashes--60~s. This latency is comparable to the other stages (data aggregation, trigger generation, alert distribution) in the real\nobreakdashes-time \ac{BNS} analysis. The 32\nobreakdashes-thread configuration is representative of how \ac{BAYESTAR} might be deployed in early Advanced \ac{LIGO}.\footnote{\ac{BAYESTAR} has been successfully ported to the Intel's Many Integrated Core~(MIC) architecture and has been tested in a 500 thread configuration on a system with dual Intel Xeon Phi 5110P coprocessors.}
